---
title: "Matching 1940 Census and SocSec Deaths: A minimal example for California"
author: "Joshua R. Goldstein"
date: "Dec. 13, 2016"
output:
  pdf_document:
    fig_width: 7
    fig_height: 9
    fig_caption: true
---


# A minimal example for doing California match

Here's an outline of what we do:

1. Read in the Census data
2. Read in the SocSec data
3. Clean data and create unique keys
4. Match and report matching rates
5. A regression to check data
6. Save results


# 0. Preliminaries

Here we load libraries and any custom functions

```{r}
library(data.table)                # we will use lots of data.table
library(readr)                     # for reading in fwf of SocSec data
library(memisc)                    # for regression table mtable()
library(stringr)                   # for str_to_upper()

# My function for getting first word of a sentence or compound name
# Much faster than stringr::word(x, 1)
get.first.word <- function(x)   
    {
        x.split <- strsplit(x, split = " ")
        x1.list <- lapply(x.split,`[`, 1) # returns NA if no name
        x1 <- unlist(x1.list)
        return(x1)
    }
```

# 1. Read in the Census data
```{r}

setwd("~/progs/ssdm")

census <- fread("~/Private/my1940/CA_clean.txt") # cleaned version has \' instead of '
# attach variable names for ca (Carl Boe's state files don't have headers)
system("head ~/Private/my1940/NORTHEAST_head.txt > tmp.txt")
names.census <- fread("tmp.txt", nrows = 10) # generates headers for this little data set
setnames(census, names(names.census))
census.orig <- census # set aside so we don't have to re-read if a problem with "census"
```

# 2. Read in the SocSec data

```{r}
## we read it in as fixed width (ssdm3 has the california geography as part of it)
tt <- readr::read_fwf("~/progs/ssdm/ssdm3",
                      fwf_widths(c(1,9, 20, 4, 15, 15, 1, 8, 8),
                                 col_names = c("mode", "ssn", "lname",
                                               "name_suffix", "fname", "mname",
                                               "vorpcode", "dod", "dob")))
socsec <- as.data.table(tt)             # convert to data.table
## restrict to CA using the SSN geography codes
socsec[, "ssn.geo" := as.numeric(substr(ssn, 1, 3))]
socsec <- socsec[ssn.geo >= 545 & ssn.geo <= 573,]
```


# 3. Clean data and create unique keys

```{r}
## A. clean the socsec data
## note: we can't restrict to Males, since no sex.
## shorten names by removing blank space at end
socsec[,lname := gsub(pattern = "\\s*$",
                  replacement = "", x = lname)]
socsec[,fname := gsub(pattern = "\\s*$",
                  replacement = "", x = fname)]
## now get birth and death year
socsec[,"byear" := as.numeric(substr(dob, 5, 9))]
socsec[,"dyear" := as.numeric(substr(dod, 5, 9))]
## birth and death month
socsec[,"bmonth" := as.numeric(substr(dob, 1, 2))]
socsec[,"dmonth" := as.numeric(substr(dob, 1, 2))]
## now get census_age (hopefully census was on April 1)
socsec[,"census_age" := ifelse(bmonth < 4,
                           1940 - byear,
                           1939 - byear)]

## B. clean the census data
census <- census.orig
census[,"fname" := str_to_upper(self_empty_name_given)]
census[,"lname" := str_to_upper(self_empty_name_surname)]
census[,"age" := as.numeric(self_residence_info_age)]
census[,"census_age" := age]
census[,"sex" := self_empty_info_gender]
census[,"fname" := get.first.word(fname)]
# income (just to take an example of one covariate)
census[,"income" := general_income,]
census[,"income" := gsub(",", "", income)]
census[,"income" := as.numeric(income)]
## HHID_NUMERIC is unique HHID for each household
## HHORDER is the order of the person in unique HH
census[,"hhid" := HHID_NUMERIC]
census[,"recno" := HHORDER]

## Keep only a minimal subset of the census variables

census <- census[,.(hhid, recno, fname, lname, age, census_age, sex,
                    income)]

dim(census)

```

## 4. Match

```{r}
## Create keys for matching, triplets of last name, first name and
## age in census
socsec[,"tmp_key" := paste0(lname, fname, census_age)]
census[,"tmp_key" := paste0(lname, fname, census_age)]
## Drop if not unique
socsec[,"n_tmp_key" := .N, by = tmp_key]
socsec.uniq <- socsec[n_tmp_key == 1,]
census[,"n_tmp_key" := .N, by = tmp_key]
census.uniq <- census[n_tmp_key == 1,]
print(nrow(socsec.uniq))
## [1] 4870219
print(nrow(census.uniq))
## [1] 6123351

## Drop if tmp_key contains weird characterrs (e.g., ??)
census.uniq <- census.uniq[!grepl("\\?", tmp_key),]
socsec.uniq <- socsec.uniq[!grepl("\\?", tmp_key),]
print(nrow(socsec.uniq))
## [1] 4870219
print(nrow(census.uniq))
## [1] 6121620, so we got rid of some (1,731 people)

## Drop if not male

census.uniq.male <- census.uniq[sex == "Male",]
print(nrow(census.uniq.male))
## Reset keys
setkey(socsec.uniq, tmp_key)
setkey(census.uniq.male, tmp_key)



## Merge on unique keys
out <- socsec.uniq[census.uniq.male]
```

# Report on sample sizes
```{r}
nrow(socsec)
nrow(socsec.uniq)
nrow(census)
nrow(census.uniq.male)                  #[1] 3061687
nrow(out)                               #[1] 3061687
## so there is one record for every uniq male in census (even those that didn't match)
out.matched <- out[!is.na(ssn),]
nrow(out.matched)
## [1] 581921
```

# 5. A regression to check data

Here we do a simple regression of income on age of death to make sure
we get something plausible.

```{r}
## we call our matched data set censoc ("[cen]sus" + "[soc]sec")
censoc <- out.matched
## create approximate age at death (without accounting for exact dates)
censoc[, age.at.death := dyear - byear]

## regress age at death on log income
censoc[, log.inc := log(income)]

m <- lm(age.at.death ~ log.inc, 
        data = censoc,
        subset = census_age >= 20 & census_age <= 40 &
            income > 0 & !is.na(income))
mtable(m)
```
So a 10 percent increase in income, increases log income by about 0.1, which would increase life expectancy by about 0.12 years.



# 6. Save results

```{r}
write.csv(out.matched, file = "ca_out_matched.csv")
```

# To do

* check names for weird characters
* check keys of those who are matched
* calculate match rates in various ways
* look at income of those matched and not-matched

